\documentclass[a4paper]{article}

\usepackage{NotesPackage2}
\usepackage{csquotes}
\usepackage{tcolorbox}
\usepackage{hyperref}

\usetikzlibrary{calc}
\usetikzlibrary{external}

\tcbuselibrary{breakable}
\tcbuselibrary{theorems}
\tcbset{highlight math style={}}

\tikzexternalize[prefix=tikz-external/]
\tikzexternaldisable

\author{Willoughby Seago}
\date{September 12, 2021}
\title{
    Methods of Theoretical Physics\\
    {\Large Vectors, Tensors and Continuum Mechanics}
}

\makeglossaries
% Add glossary entries here
\newacronym{eg}{EG}{Example}  % this is an example

\newcommand{\notesVersion}{1.1}
\newcommand{\notesDate}{07/03/2021}

\newcommand{\nxmMatrices}[3]{\mathcal{M}_{{#1}\times{#2}}({#3})}
\renewcommand{\ident}{I}
\newcommand{\swap}[2]{{#1} \leftrightarrow {#2}}
\newcommand{\swapEquals}[2]{\overset{\swap{#1}{#2}}{=}}
\newcommand{\generalLinearGroup}{\mathrm{GL}}
\newcommand{\orthogonalGroup}{\mathrm{O}}
\newcommand{\specialOrthogonalGroup}{\mathrm{SO}}
\def\centerarc[#1](#2)(#3:#4:#5)% Syntax: [draw options] (center) (initial angle:final angle:radius)
{ \draw[#1] ($(#2)+({#5*cos(#3)},{#5*sin(#3)})$) arc (#3:#4:#5); }
\newcommand{\lieAlgebra}[1]{\mathfrak{#1}}
\newcommand{\basis}{\mathcal{B}}
\newcommand{\veprime}[1]{\vv{e'_{#1}}}
\newcommand{\vepprime}[1]{\vv{e''_{#1}}}
\newcommand{\nindices}[2]{{ {#1}_1 {#1}_2 \dotsm {#1}_{#2} }}
\DeclareMathOperator{\GRAD}{grad}
\DeclareMathOperator{\DIV}{div}
\DeclareMathOperator{\CURL}{curl}
\newcommand{\vd}[1]{\vv{\mathrm{d}{#1}}}
\newcommand{\freeEnergy}{\mathcal{F}}

\includeonly{parts/vectors_and_tensors}

\begin{document}
    \pagenumbering{roman}  % Number contents pages and glossaries with roman numerals
    \maketitle
    These are my notes for the \textit{vectors, tensors, and continuum mechanics} part of the \textit{methods of theoretical physics} course from the University of Edinburgh as part of the third year of the theoretical physics degree.
    When I took this course in the 2020/21 academic year it was taught by Dr Mehdi Bouzid\footnote{\url{https://www.ph.ed.ac.uk/people/mehdi-bouzid}}.
    These notes are based on the lectures delivered as part of this course and the notes provided as part of this course which were originally written by Professor Alexander Morozov\footnote{\url{https://ww.ph.ed.ac.uk/people/alexander-morozov}}.
    The content within is correct to the best of my knowledge but if you find a mistake or just disagree with something or think it could be improved please let me know.
    
    These notes were produced using \LaTeX\footnote{\url{https://www.latex-project.org/}}.
    Graphs where plotted using Matplotlib\footnote{\url{https://matplotlib.org/}}, NumPy\footnote{\url{https://numpy.org/}}, and SciPy\footnote{\url{https://scipy.org/scipylib/}}.
    Diagrams were drawn with tikz\footnote{\url{https://www.ctan.org/pkg/pgf}}.
    
    This is version \notesVersion~of these notes, which is up to date as of \notesDate.
    \begin{flushright}
        Willoughby Seago
        
        s1824487@ed.ac.uk
    \end{flushright}
    \clearpage
    \tableofcontents
%    \listoftheorems[ignoreall, show={theorem,corollary,lemma}]
%    \renewcommand{\listtheoremname}{List of Definitions}
%    \listoftheorems[ignoreall, show=definition]
    \printglossary[type=\acronymtype, title=Acronyms]
    \pagenumbering{arabic}  % Number rest of document with numbers
    \clearpage
    \begingroup
    \let\clearpage\relax  % "\begingroup, \let\clearpage\relax, \endgroup" stops automatic pagebreaks after each include
    \include{parts/vectors_and_tensors}
    \endgroup
    
    
    \part{Linear Elasticity}
    \section{The Deformation Tensor}
    \subsection{Deformation Types}
    There are many ways that a body can be deformed.
    Some of the most common are shears, compressions, and stretches.
    \begin{figure}[ht]
        \centering
        \tikzsetnextfilename{elementary-deformations}
        \begin{tikzpicture}
            \tikzset{body/.style={fill=lightgray, fill opacity=0.3, thick}}
            \pgfmathsetmacro{\boxHeight}{2}
            \pgfmathsetmacro{\boxLength}{3}
            % Coordinate naming: (b/t|f/b|l/r) corresponds to corner (base/top|front/back|left/right)
            \coordinate (bfl) at (0, 0);
            \coordinate (bbl) at (0.5, 0.5);
            \coordinate (bfr) at ($(bfl) + (\boxLength, 0)$);
            \coordinate (bbr) at ($(bbl) + (\boxLength, 0) - (1, 0)$);
            \coordinate (tfl) at ($(bfl) + (0, \boxHeight)$);
            \coordinate (tbl) at ($(bbl) + (0, \boxHeight) - (0, 1)$);
            \coordinate (tbr) at ($(tbl) + (\boxLength, 0) - (1, 0)$);
            \coordinate (tfr) at ($(tfl) + (\boxLength, 0)$);
            \draw[body] (bfl) -- (bfr) -- (tfr) -- (tfl) -- cycle; % Front 
            \draw[body] (bbl) -- (bbr) -- (tbr) -- (tbl) -- cycle; % Back 
            \draw[body] (bfl) -- (bbl) -- (tbl) -- (tfl) -- cycle; % Left 
            \draw[body] (bfr) -- (bbr) -- (tbr) -- (tfr) -- cycle; % Right
            \draw[body] (bfl) -- (bfr) -- (bbr) -- (bbl) -- cycle; % Base
            \draw[body] (tfl) -- (tfr) -- (tbr) -- (tbl) -- cycle; % Top
            \node at ($(bfl)!0.5!(bfr) - (0 ,1)$) {Undeformed};
            
            \begin{scope}[xshift=5cm]
                \coordinate (bfl) at (0, 0);
                \coordinate (bbl) at (0.5, 0.5);
                \coordinate (bfr) at (3, 0);
                \coordinate (bbr) at (3, 0.5);
                \coordinate (tfl) at (1, 2);
                \coordinate (tbl) at (1, 1.5);
                \coordinate (tbr) at (3.5, 1.5);
                \coordinate (tfr) at (4, 2);
                \draw[body] (bfl) -- (bfr) -- (tfr) -- (tfl) -- cycle; % Front 
                \draw[body] (bbl) -- (bbr) -- (tbr) -- (tbl) -- cycle; % Back 
                \draw[body] (bfl) -- (bbl) -- (tbl) -- (tfl) -- cycle; % Left 
                \draw[body] (bfr) -- (bbr) -- (tbr) -- (tfr) -- cycle; % Right
                \draw[body] (bfl) -- (bfr) -- (bbr) -- (bbl) -- cycle; % Base
                \draw[body] (tfl) -- (tfr) -- (tbr) -- (tbl) -- cycle; % Top
                \node at ($(bfl)!0.5!(bfr) - (0 ,1)$) {Shear};
            \end{scope}
            
            \begin{scope}[yshift=-4cm]
                \pgfmathsetmacro{\boxLength}{2}
                \coordinate (bfl) at (0, 0);
                \coordinate (bbl) at (0.5, 0.5);
                \coordinate (bfr) at ($(bfl) + (\boxLength, 0)$);
                \coordinate (bbr) at ($(bbl) + (\boxLength, 0) - (1, 0)$);
                \coordinate (tfl) at ($(bfl) + (0, \boxHeight)$);
                \coordinate (tbl) at ($(bbl) + (0, \boxHeight) - (0, 1)$);
                \coordinate (tbr) at ($(tbl) + (\boxLength, 0) - (1, 0)$);
                \coordinate (tfr) at ($(tfl) + (\boxLength, 0)$);
                \draw[body] (bfl) -- (bfr) -- (tfr) -- (tfl) -- cycle; % Front 
                \draw[body] (bbl) -- (bbr) -- (tbr) -- (tbl) -- cycle; % Back 
                \draw[body] (bfl) -- (bbl) -- (tbl) -- (tfl) -- cycle; % Left 
                \draw[body] (bfr) -- (bbr) -- (tbr) -- (tfr) -- cycle; % Right
                \draw[body] (bfl) -- (bfr) -- (bbr) -- (bbl) -- cycle; % Base
                \draw[body] (tfl) -- (tfr) -- (tbr) -- (tbl) -- cycle; % Top
                \node at ($(bfl)!0.5!(bfr) - (0 ,1)$) {Compression};
            \end{scope}
        
            \begin{scope}[yshift=-4cm, xshift=5cm]
                \pgfmathsetmacro{\boxLength}{4}
                \coordinate (bfl) at (0, 0);
                \coordinate (bbl) at (0.5, 0.5);
                \coordinate (bfr) at ($(bfl) + (\boxLength, 0)$);
                \coordinate (bbr) at ($(bbl) + (\boxLength, 0) - (1, 0)$);
                \coordinate (tfl) at ($(bfl) + (0, \boxHeight)$);
                \coordinate (tbl) at ($(bbl) + (0, \boxHeight) - (0, 1)$);
                \coordinate (tbr) at ($(tbl) + (\boxLength, 0) - (1, 0)$);
                \coordinate (tfr) at ($(tfl) + (\boxLength, 0)$);
                \draw[body] (bfl) -- (bfr) -- (tfr) -- (tfl) -- cycle; % Front 
                \draw[body] (bbl) -- (bbr) -- (tbr) -- (tbl) -- cycle; % Back 
                \draw[body] (bfl) -- (bbl) -- (tbl) -- (tfl) -- cycle; % Left 
                \draw[body] (bfr) -- (bbr) -- (tbr) -- (tfr) -- cycle; % Right
                \draw[body] (bfl) -- (bfr) -- (bbr) -- (bbl) -- cycle; % Base
                \draw[body] (tfl) -- (tfr) -- (tbr) -- (tbl) -- cycle; % Top
                \node at ($(bfl)!0.5!(bfr) - (0 ,1)$) {Stretch};
            \end{scope}
        \end{tikzpicture}
        \caption{Common deformations}
        \label{fig:common deformations}
    \end{figure}
    These are shown in figure~\ref{fig:common deformations}.
    Very rarely does only one of these deformations occur.
    For example if you compress an object along one axis the matter has to go somewhere so it tends to increase in size along the other axes.
    We aim to come up with a way to describe a general deformation.
    In this course we are interested in linear elasticity which means that we assume all deformations are sufficiently small that things can safely be expanded only up to linear order in the relevant variable.
    
    \subsection{Geometry of Deformations}
    A body in equilibrium has its position, orientation, and shape set by the positions, \(\{\vv{r}\}\), of its material points.
    What counts as a material point exactly isn't important, its simply a point that we can track as a body moves and deforms, it could be an atom or molecule or it might be a region of material.
    
    Suppose the body is then deformed in such a way that a material point that was at \(\vv{r}\) before the deformation is now at \(\vv{r'}\).
    We define the \define{displacement vector}, \(\vv{u}\), as the displacement of the point under this deformation:
    \[\vv{u} = \vv{r} - \vv{r'}.\]
    In a translation or rotation the shape of the body is fixed and therefore \(\vv{u}\) is the same for all points.
    In a deformation however the shape of the body can change and therefore \(\vv{u}\) is not the same for all points.
    We can think of \(\vv{u}\) as a vector field mapping points to there deformed points.
    This introduces a double dependence on \(\vv{r}\): first \(\vv{r}\) is the point that is being moved and second \(\vv{r}\) is the point at which we are evaluating \(\vv{u}(\vv{r})\).
    By simply rearranging the definition of \(\vv{u}\) we have \(\vv{r'} = \vv{r} + \vv{u}(\vv{r})\) and \(\{\vv{r'}\}\) gives us the shape, orientation, and position of the deformed body.
    
    Consider two points, \(\vv{r}\) and \(\vv{r} + \vd{r}\), on the undeformed body.
    The Cartesian coordinates of these points are then \(x_i\) and \(x_i + \dd{x_i}\) respectively.
    After the deformation we have
    \begin{align*}
        x_i&\mapsto x_i + u_i(\vv{r}),\\
        x_i + \dd{x_i}&\mapsto x_i + \dd{x_i} + u_i(\vv{r} + \vd{r}).
    \end{align*}
    The distance between the two points before the deformation is \(\dd{l}\), which by Pythagoras' theorem is simply
    \[\dd{l}^2 = \dd{x_1}^2 + \dd{x_2}^2 + \dd{x_3}^2 = \dd{x_i}^2.\]
    Note that even though the \(i\) only appears once since it appears in a squared term we view it as \(\dd{x_i}\dd{x_i}\) so there is an implied summation.
    The distance between the two points after the deformation is \(\dd{l'}\).
    To calculate this we similarly need to sum the square of the differences between the coordinates:
    \begin{align*}
        \dd{l'}^2 &= [(x_i + \dd{x_i} + u_i(\vv{r} + \vd{r})) - (x_i + u_i(\vv{r}))]^2\\
        &= [\dd{x_i} + u_i(\vv{r} + \vd{r}) - u_i(\vv{r})]^2.
    \end{align*}
    We can Taylor expand \(u_i\) about \(\vv{r}\) to first order as
    \[u_i(\vv{r} + \vd{r}) = u_i(\vv{r}) + \pdv{u_i}{x_k}\dd{x_k} + \order{\dd{x_k}^2} \implies u_i(\vv{r} + \vd{r}) - u_i(\vv{r}) \approx \pdv{u_i}{x_k}\dd{x_k}.\]
    Using this we have
    \begin{align*}
        \dd{l'}^2 \approx \left[\dd{x_i} + \pdv{u_i}{x_k}\dd{x_k}\right]^2\\
        &= \dd{x_i}^2 + 2\pdv{u_i}{x_k}\dd{x_i}\dd{x_k} + \pdv{u_i}{x_k}\pdv{u_i}{x_j}\dd{x_k}\dd{x_j}.
    \end{align*}
    The first term is simply \(\dd{l}^2\).
    The second term can be expanded as
    \begin{align*}
        2\pdv{u_i}{x_k}\dd{x_i}\dd{x_k} &= \pdv{u_i}{x_k}\dd{x_i}\dd{x_k} + \pdv{u_i}{x_k}\dd{x_i}\dd{x_k}
        \shortintertext{renaming \(\swap{i}{k}\) in the second term only we have}
        &= \pdv{u_i}{x_k}\dd{x_i}\dd{x_k} + \pdv{u_k}{x_k}\dd{x_k}\dd{x_i}\\
        &= \pdv{u_i}{x_k}\dd{x_i}\dd{x_k} + \pdv{u_k}{x_k}\dd{x_i}\dd{x_k}.
    \end{align*}
    For the final term if we rename \(i\rightarrow l\) and \(j\to i\) we have
    \[\pdv{u_i}{x_k}\pdv{u_i}{x_j}\dd{x_k}\dd{x_j} = \pdv{u_l}{x_k}\pdv{u_l}{x_k}\dd{x_k}\dd{x_k}\]
    Putting this into \(\dd{l'}^2\) we have
    \begin{align*}
        \dd{l'}^2 &= \dd{l}^2 + \left(\pdv{u_i}{x_k} + \pdv{u_k}{x_i}\right) \dd{x_i}\dd{x_k} + \pdv{u_l}{x_i}\pdv{u_l}{x_k}\\
        &= \dd{l}^2 + \left[\pdv{u_i}{x_k} + \pdv{u_k}{x_i} + \pdv{u_l}{x_i}\pdv{u_l}{x_k}\right]\dd{x_i}\dd{x_k}\\
        &= \dd{l}^2 + 2u_{ik}\dd{x_i}\dd{x_k}
    \end{align*}
    where
    \[\tcbhighmath{
        u_{ik} = \frac{1}{2}\left(\pdv{u_i}{x_k} + \pdv{u_k}{x_i} + \pdv{u_l}{x_k}\pdv{u_l}{x_k}\right)
    }\]
    are the coordinates of a rank 2 Cartesian tensor, \(u\), called the \define{deformation tensor} or the \define{strain tensor}.
    Note that the factor of \(1/2\) and therefore factor of 2 in \(\dd{l'}^2\) are simply a matter of convention.
    Note that sometimes the last term is dropped as it is of higher order so not as important for small deformations.
    The notation here can be a bit confusing as we use \(u\) for the displacement vector and the deformation tensor.
    The two uses can be distinguished by the rank of the object in question.
    
    The components, \(u_{ik}\), are symmetric in \(i\) and \(k\) (\(u_{ik} = u_{ki}\)).
    This means that there are only 6 free components and the other three are fixed by symmetry.
    One corollary of this si that \(u\) can always be diagonalised.
    Let \(u^{(i)}\) be the eigenvalues, or principle components, of \(u\).
    Then there in the eigenbasis, \(\{\vv{u^{(i)}}\}\), also known as the principle axes, we have
    \[
        u = 
        \begin{pmatrix}
            u^{(1)} & 0 & 0\\
            0 & u^{(2)} & 0\\
            0 & 0 & u^{(3)}
        \end{pmatrix}
        .
    \]
    Suppose we work in the eigenbasis.
    Then
    \begin{align*}
        \dd{l'}^2 &= \dd{l}^2 + 2u_{ik}\dd{x_i}\dd{x_k}\\
        &= \dd{x_i}^2 + 2u_{ik}\dd{x_i}\dd{x_k}\\
        &= \delta_{ik}\dd{x_i}\dd{x_k} + 2u_{ik}\dd{x_i}\dd{x_k}\\
        &= (\delta_{ik} + 2u_{ik})\dd{x_i}\dd{x_k}\\
        &= (1 + 2u_{11})\dd{x_1}^2 + (1 + 2u_{22})\dd{x_2}^2 + (1 + 2u_{33})\dd{x_3}^2\\
        &= (1 + 2u^{(1)})\dd{x_1}^2 + (1 + 2u^{(2)})\dd{x_2}^2 + (1 + 2u^{(3)})\dd{x_3}^2
    \end{align*}
    So the proportional change of separation along each axis is
    \[\dd{xi'} = \sqrt{1 + 2u^{(i)}}\dd{x_i}\]
    and hence
    \[\frac{\dd{x_i'} = \dd{x_i}}{\dd{x_i}} = \sqrt{1 + 2u^{(i)}} - 1.\]
    Since the deformation is small \(u^{(i)}\) must be small and so we expand this in terms of \(u^{(i)}\):
    \[\sqrt{1 + 2u^{(i)}} = (1 + 2u^{(i)})^{1/2} \approx 1 + u^{(i)} + \order{u^{(i)}{^2}}\]
    so we have
    \[\dd{x_i'} \approx (1 + u^{(i)})\dd{x_i}, \qquad\text{and}\qquad \frac{\dd{x_i'} - \dd{x_i}}{\dd{x_i}} \approx u^{(i)}.\]
    
    Suppose we wish to know the volume change.
    Consider a small volume, \(\dd{V} = \dd{x_1}\dd{x_2}\dd{x_3}\), in the undeformed body.
    After the deformation this becomes
    \begin{align*}
        \dd{V'} &= (1 + u^{(1)})(1 + u^{(2)})(1 + u^{(3)})\dd{x_1}\dd{x_2}\dd{x_3}\\
        &\approx (1 + u^{(1)} + u^{(2)} + u^{(3)})\dd{V}\\
        &= [1 + \Tr (u)]\dd{V}
    \end{align*}
    where we have dropped any terms of second order or higher.
    Hence we have
    \[\frac{\dd{V'} - \dd{V}}{\dd{V}} = \Tr u.\]
    This is a useful result because the trace is invariant under a coordinate transform, this can be shown by considering a transformation \(L\in\specialOrthogonalGroup(3)\):
    \[A'_{ii} = l_{i\alpha}l_{i\beta}A_{\alpha\beta} = L\trans_{\alpha i}L_{i\beta}A_{\alpha\beta} = (L\trans L)_{\alpha\beta} A_{\alpha\beta} = \delta_{\alpha\beta}A_{\alpha\beta} = A_{\alpha\alpha} = \Tr(A).\]
    \begin{example}
        \textit{Consider a cuboid that is stretched along the \(x_1\) axis so the length increases from \(l\) to \(l + \Delta l\) for some fractional change, \(\alpha\). Suppose that this also causes the cuboid to contract along the other two axes by some amount such that the fractional change, \(\beta\), is the same for both axes. What is the deformation tensor for this deformation?}
        
        First consider the \(x_1\) axis:
        \[\frac{\Delta l}{l} = \frac{x_i' - x_i}{x_i} = \frac{u^{(1)}}{x_i} = \alpha.\]
        So \(u^{(1)} = \alpha x_1\).
        Now consider the \(x_2\) axis.
        If we think of this as the height, \(h\), of the cuboid then we have
        \[\frac{\Delta h}{h} = \frac{x_2' - x_2}{x_2} = \frac{u^{(2)}}{x_2} = -\beta.\]
        So \(u^{(2)} = -\beta x_2\), notice that this is negative since the cuboid is compressed along this axis.
        Similarly if we think about the \(x_3\) axis as width, \(w\), of the cuboid we have
        \[\frac{\Delta h}{h} = \frac{x_3' - x_3}{x_3} = \frac{u^{(3)}}{x_3} = -\beta.\]
        So \(u^{(3)} = -\beta x_3\).
        Hence the deformation tensor in the principle axis basis has components
        \[
            u =
            \begin{pmatrix}
                \alpha & 0 & 0\\
                0 & -\beta & 0\\
                0 & 0 & -\beta
            \end{pmatrix}
            .
        \]
    \end{example}

    \section{The Stress Tensor}
    An undeformed body is in equilibrium with its surroundings.
    This means that all parts of the body are in equilibrium with each other.
    The total force and total torque the body are zero.
    When the body is deformed by external forces it is moved away from equilibrium creating internal forces in the body.
    
    Consider a small portion of the body.
    Any forces acting on this portion must have a net effect of acting only at the surface as forces within the body cancel out.
    Let \(\vv{F}\) be the force per unit volume, or force density, acting on the volume.
    The total force on the volume is simply
    \[\int_{V} \vv{F} \dd{V}.\]
    Since the force acts only at the surface there exists a tensor, \(\sigma_{ij}\), such that we can write the total force as a surface integral,
    \begin{equation}\label{eqn:int F dV = int sigma dA}
        \int_V F_i \dd{V} = \int_{\partial V} \sigma_{ij} \dd{A_j},
    \end{equation}
    where \(\partial V\) is the boundary of the volume and \(\dd{A_j} = n_j\dd{A}\) is an area element along the surface normal \(\vh{n}\).
    By the divergence theorem we can then write
    \[\int_{\partial V}\sigma_{ij}\dd{A_j} = \int_{V}\pdv{\sigma_{ij}}{x_j} \dd{V}.\]
    Hence
    \[\tcbhighmath{
        F_i = \pdv{\sigma_{ij}}{x_j}.
    }\]
    In conclusion if the only forces acting on a body act on its surface then the internal forces in the body can be written as the divergence of a second rank tensor, \(\sigma\), which is called the \define{stress tensor}.
    
    For a physical understanding of what \(\sigma\) represents consider equation~\ref{eqn:int F dV = int sigma dA}.
    This gives us
    \[\dd{F_i} = \sigma_{ij}\dd{A_j} = \sigma_{ij}n_j\dd{A}\]
    where \(\dd{F_i} = F_i\dd{V}\) is the force on the volume element \(\dd{V}\) in the direction \(x_i\).
    From this we can see that \(\sigma_{ij}\) is the force per unit area in the direction \(x_i\) acting on the surface element normal to \(x_j\).
    This can be seen more clearly in figure~\ref{fig:stress tensor} which shows the stress tensor on a cube.
    Each component of \(\sigma_{i3}\) is proportional to the force acting on the face normal to \(x_3\) and the force acts in the \(x_i\) direction.
    \begin{figure}[ht]
        \centering
        \tikzsetnextfilename{stress-tensor}
        \begin{tikzpicture}
            \tikzset{body/.style={gray, very thick, fill=lightgray!50, fill opacity=0.5}}
            \tikzset{tensor/.style={ultra thick, ->, >=latex}}
            \draw[body] (1, 1) rectangle (5, 5); % back
            \draw[body] (0, 0) -- (1, 1) -- (1, 5) -- (0, 4) -- cycle; % left
            \draw[body] (0, 0) -- (1, 1) -- (5, 1) -- (4, 0) -- cycle; % bottom
            \draw[body] (4, 0) -- (5, 1) -- (5, 5) -- (4, 4) -- cycle; % right
            \draw[body] (0, 4) -- (1, 5) -- (5, 5) -- (4, 4) -- cycle; % top
            \draw[body] (0, 0) rectangle (4, 4); % front
            \draw[tensor] (2.25, 4.5) -- (2.25, 6) node[above] {\(\sigma_{33}\)};
            \draw[tensor] (2.25, 4.5) -- (3.75, 4.5) node[right] {\(\sigma_{23}\)};
            \draw[tensor] (2.25, 4.5) -- ($(2.25, 4.5)!0.8!(1.18933, 3.439)$) node[below] {\(\sigma_{13}\)};
            \begin{scope}[xshift=1cm, yshift=1cm]
                \draw[->, >=latex] (6, 2) -- (6, 3.5) node[above] {\(x_3\)};
                \draw[->, >=latex] (6, 2) -- (7.5, 2) node[right] {\(x_2\)};
                \draw[->, >=latex] (6, 2) -- ($(6, 1)!0.8!(4.939, 0.939)$) node[below] {\(x_1\)};
            \end{scope}
            \node at (-3.5, 0) {}; % make the cube centred
        \end{tikzpicture}
        \caption{The stress tensor}
        \label{fig:stress tensor}
    \end{figure}
    
    Now consider the torque that acts on the volume as a result of the neighbouring volume elements inside the body.
    This is given by
    \begin{align*}
        \left[ \int \vv{r}\times\vv{F} \dd{V} \right]_i &= \int \varepsilon_{ijk} x_jF_k \dd{V}\\
        &=  \int (x_2F_3 - x_3F_2)\delta_{i1} \dd{V} + \int (x_3F_1 - x_1F_3)\delta_{i2} \dd{V} + \int (x_1F_2 - x_2F_1)\delta_{i3} \dd{V}.
    \end{align*}
    Consider just one of these terms:
    \begin{align*}
        \int (x_iF_j - x_jF_i) \dd{V} &= \int \left( x_i\pdv{\sigma_{jk}}{x_k} - x_j\pdv{\sigma_{ik}}{x_k} \right)\dd{V}\\
        &= \int \pdv{x_k} (x_i\sigma_{ik} - x_j\sigma_{ik}) \dd{V} - \int \left( \pdv{x_i}{x_k}\sigma_{jk} - \pdv{x_j}{x_k}\sigma_{ik} \right)\dd{V}\\
        &= \int (x_i\sigma_{jk} - x_j\sigma_{ik})\dd{A_k} - \int (\delta_{ik}\sigma_{jk} - \delta_{jk}\sigma_{ik})\dd{V}\\
        &= \int (x_i\sigma_{jk} - x_j\sigma_{ik})\dd{A_k} - \int (\sigma_{ji} - \sigma_{ij})\dd{V}.\\
    \end{align*}
    Here we have again used the divergence theorem to replace a volume integral with a surface integral.
    The same logic that we applied to the forces here leads us to the conclusion that the torque must be able to be written as a surface integral.
    Therefore we require the second term to be zero.
    Since this must be the case for an arbitrary volume element we have
    \[\sigma_{ij} = \sigma_{ji}\]
    so the stress tensor is symmetric.
    
    In the absence of external forces there is no net force applied to any volume element and so
    \[F_i = \pdv{\sigma_{ij}}{x_j} = 0 \iff \DIV\sigma = 0.\]
    If there is an external force that acts on all volume elements then
    \[\pdv{\sigma_{ij}}{x_j} + f_i = 0\]
    where \(f_i\) is the force per unit volume in the \(x_i\) direction.
    For example if a body experiences gravity along the \(x_2\) axis then we have
    \[\pdv{\sigma_{1j}}{x_j} = \pdv{\sigma_{2j}}{x_j} - mg = \pdv{\sigma_{3j}}{x_j} = 0.\]
    The boundary equations for this differential equation come from specifying the force applied to each surface of the body.
    For each surface we must specify the force, \(P_i\), applied to area \(A\) along \(x_j\) which is given by \(P_i = \sigma_{ij}n_jA\) where \(\vh{n}\) is the surface normal.
    
    \begin{example}
        \textit{Consider a cylinder stretched by two equal and opposite forces applied to its ends. What is the stress tensor?}
        \begin{figure}[ht]
            \centering
            \tikzsetnextfilename{cylinder-stetch-stress-example}
            \begin{tikzpicture}
                \tikzset{body/.style={gray, very thick, fill=lightgray!25}}
                \tikzset{force/.style={ultra thick, ->, >=latex}}
                \draw[body] (0, 0) circle [x radius=0.5cm, y radius=1cm];
                \fill[body] (0, 1) rectangle (5, -1);
                \draw[body] (0, -1) rectangle (5, -1);
                \draw[body] (0, 1) rectangle (5, 1);
                \draw[body] (5, 0) circle [x radius=0.5cm, y radius=1cm];
                \draw[force] (5, 0) -- (7, 0) node[right] {\(P\)};
                \draw[force] (-0.5, 0) -- (-2, 0) node[left] {\(-P\)};
            \end{tikzpicture}
            \caption{Cylinder being stretched.}
            \label{fig:example 1 stress tensor}
        \end{figure}
        
        Figure~\ref{fig:example 1 stress tensor} shows the setup.
        Let \(a\) be the end with force \(-P\) per unit area and \(b\) be the end with force \(P\) per unit area.
        Let the \(x_1\)-axis run from \(a\) to \(b\), the \(x_2\)-axis be into the page and the \(x_3\) axis be up.
        Then the boundary conditions are as follows:
        \begin{itemize}
            \item At \(b\) the outer normal is \(\ve{1}\) and so \(\sigma_{ij}(b)\delta_{j1}A = P\delta_{i1}\) where \(A\) is the area of the circle.
            \item At \(a\) the outer normal is \(-\ve{1}\) and so \(-\sigma_{ij}(a)\delta_{j1}A = -P\delta_{i1}\).
            \item On the curved surface of the cylinder \(\vh{n}\) is perpendicular to \(\ve{1}\) and so \(\vh{n} = \alpha \ve{2} + \beta\ve{3}\).
            Since no force acts in this direction we have \(\sigma_{ij}n_j\tilde{A} = 0\) where \(\tilde{A}\) is the area of the element of the curved surface that we are currently considering.
            Expanding these we have
            \begin{align*}
                \alpha\sigma_{12} + \beta\sigma_{13} &= 0,\\
                \alpha\sigma_{22} + \beta\sigma_{23} &= 0,\\
                \alpha\sigma_{32} + \beta\sigma_{33} &= 0,
            \end{align*}
        \end{itemize}
        Since, in general, there are surface elements where \(\alpha, \beta \ne 0\) we must have \(\sigma_{ij} = 0\) for \(i, j \ne 1\).
        Hence
        \[
            \sigma_{ij} = 
            \begin{pmatrix}
                P/A & 0 & 0\\
                0 & 0 & 0\\
                0 & 0 & 0
            \end{pmatrix}
            .
        \]
        
    \end{example}
    
    \begin{example}\label{exa:isotropic compression stress tensor}
        \textit{What is the stress tensor for an isotropic compression?}
        
        An isotropic compression involves the same force on each face.
        Let \(p\) is the force per unit area.
        For a compression \(p\) points into the surface, i.e. anti-parallel to the surface normal.
        Also the only force on a face is perpendicular to that face so the off diagonal elements of the stress tensor are zero.
        Hence \(\sigma_{ij}n_jA = -p\delta_{ik}n_kA = -p\ident\).
        \[
            \sigma_{ij} = 
            \begin{pmatrix}
                -p & 0 & 0\\
                0 & -p & 0\\
                0 & 0 & -p
            \end{pmatrix}
            .
        \]
    \end{example}
    \begin{example}\label{exa:shear stress tensor}
        \textit{What is the stress tensor for a shear of the surface normal to \(x_2\) along the \(x_1\) direction?}
        
        The only force is applied to the surface with surface normal \(\ve{2}\).
        If the force is \(F\) and it is in the \(x_1\) direction then we must have \(\sigma_{12} = F\).
        Since the stress tensor is symmetric we must also have \(\sigma_{21} = F\).
        Hence
        \[
            \sigma_{ij} =
            \begin{pmatrix}
                0 & F & 0\\
                F & 0 & 0\\
                0 & 0 & 0
            \end{pmatrix}
            .
        \]
    \end{example}
    
    \subsection{Relation Between Deformation and Stress Tensors}
    Suppose we have a medium being deformed under some external force so that the stress tensor is \(\sigma\) and the deformation tensor is \(u\).
    We want to find a relationship between \(\sigma\) and \(u\).
    Imagine we perform some extra work on the system changing the displacement vector \(\vv{u}\) to \(\vv{u} + \vv{\delta u}\) (\(u_i\) to \(u_i + \delta u_i\)).
    We are free to choose how to do this work and so we choose to do it in a way such that \(\delta u_i\) vanishes at the surface.
    The work done by internal forces is given by
    \[\int \delta W\dd{V} = \int \vv{F}\cdot\vv{\delta u}\dd{V}\]
    where \(\delta W\) is the work done on a single volume element and \(\vv{F}\) is the internal force that opposes the deformation.
    By the definition of the stress tensor we have
    \[\int \delta W \dd{V} = \int \pdv{\sigma_{ik}}{x_k}\delta u_i \dd{V} = \int \pdv{x_k}(\delta u_i \sigma_{ik}) \dd{V} - \int \sigma_{ik}\pdv{x_k}\delta u_i.\]
    We can write the first integral as a surface integral using the divergence theorem.
    We can write the second integral in a symmetric form.
    \[\int \delta W \dd{V} = \int \sigma_{ik}\delta u_i n_k \dd{A} - \frac{1}{2} \int \left( \sigma_{ik}\pdv{x_k}\delta u_i + \sigma_{ki}\pdv{x_i}\delta u_k \right)\dd{V}.\]
    Since we chose \(\delta u_i\) to vanish at the surface the first integral vanishes.
    For the second integral we exploit the symmetry of the stress tensor:
    \[\int \delta W \dd{V} = -\frac{1}{2}\int \sigma_{ik}\left( \pdv{x_k}\delta u_i + \pdv{x_i}\delta u_k \right) \dd{V} = -\int \sigma_{ik}\delta u_{ik} \dd{V}\]
    where
    \[\delta u_{ik} = \frac{1}{2}\left( \pdv{u_i}{x_k} + \pdv{u_k}{x_i} \right).\]
    Notice that this is simply the deformation tensor after dropping terms of second order.
    
    \section{Stress and Strain}
    Recall that in the last section we derived the result
    \[\int\delta W\dd{V} = -\int\sigma_{ik}\delta u_{ik}\dd{V}.\]
    The free energy is defined as
    \[\freeEnergy = U - TS\]
    where \(U\) is the internal energy, \(T\) is the temperature, and \(S\) is the entropy.
    The value derived above is the work done, which is equal to the negative of the internal energy.
    The infinitesimal free energy change due to the deformation is
    \[\dd{\freeEnergy} = \dd{(U - TS)} = \dd{U} - S\dd{T}\]
    using \(\dd{U} = -\delta W\) we have
    \[\dd{\freeEnergy} = -\delta W - S\dd{T} = \sigma_{ik}\delta u_{ik} - S\dd{T}.\]
    At constant temperature we have \(\dd{T} = 0\) and so
    \[\dd{\freeEnergy} = \sigma_{ik}\delta u_{ik}.\]
    Hence
    \[\sigma_{ik} = \pdvconst{\freeEnergy}{u_{ik}}{T}.\]
    Hence \(\freeEnergy\) is a function of the deformation tensor, \(u\).
    For a homogenous, isotropic medium, which is the only sort we are interested in here, there is no preferred direction and hence \(\freeEnergy\) can only depend on quantities that are invariant under rotations.
    These quantities are \(\Tr u = u_{ii}\), \(u_{ik}u_{ki}\), and \(u_{ij}u_{jk}u_{ki}\).
    All higher order tensor invariants can be written in terms of these quantities.
    Further we consider only small deformations and therefore neglect any terms that appear higher than second order.
    This leaves us with a generalised version of Hooke's law, which states that the energy stored in a spring with spring constant \(k\) is \(kx^2/2\) where \(x\) is the displacement of the spring from equilibrium.
    We can think of this simply as Taylor expanding \(\freeEnergy\) in terms of \(u_{ik}\).
    Finally notice that two deformations with \(u_{ij}\) and \(-u_{ij}\) produce the same free energy change and hence \(u\) can only appear in even powers since the transformation \(u \to -u\) cannot be allowed to introduce a negative.
    Hence the free energy is given by
    \[\freeEnergy = \frac{\lambda}{2}u_{ii}^2 + \mu u_{ik}^2\]
    where we have used the fact that \(u\) is symmetric so \(u_{ik}u_{ki} = u_{ik}u_{ik} = u_{ik}^2\), note that we still sum over \(i\) and \(k\).
    Here the constants of proportionality are \(\lambda/2\) and \(\mu\).
    \(\lambda\) and \(\mu\) are called the Lam\`e coefficients.
    
    We saw in examples~\ref{exa:isotropic compression stress tensor} and \ref{exa:shear stress tensor} that for an isotropic compression the stress tensor is diagonal.
    Similarly for a shear the stress tensor will have zeros on the diagonal.
    Using this we can split any stress tensor into two parts.
    A diagonal tensor for the compression component and a tensor with only non-zero elements off the diagonal for the shear component.
    
    Now consider differentials of the free energy:
    \begin{align*}
        \dd{\freeEnergy} = \frac{\lambda}{2}\dd{(u_{ll}^2)} + 2\mu\dd{u_{ik}^2}\\
        &= \lambda u_{ll}\dd{u_{ll}} + 2\mu u_{ik}\dd{u_{ik}}
    \end{align*}
    Consider the differential of the trace:
    \[\dd{u_{ll}} = \pdv{u_{ll}}{u_{ik}}\dd{u_{ik}}.\]
    The derivative can be computed easily by first noting that
    \[\pdv{u_{ab}}{u_{ik}} = \delta_{ia}\delta_{kb}\]
    and so
    \[\pdv{u_{ll}}{u_{ik}} = \delta_{il}\delta_{kl} = \delta_{ik}.\]
    Hence
    \[\dd{u_{ll}} = \delta_{ik}\dd{u_{ik}}.\]
    Substituting this into the differential of the free energy we have
    \[\dd{\freeEnergy} = \lambda u_{ll}\delta_{ik}\dd{u_{ik}} + 2\mu u_{ik}\dd{u_{ik}}.\]
    Dividing through by \(\dd{u_{ik}}\) and assuming convergence etc. we have
    \[\sigma_{ik} = \pdv{\freeEnergy}{u_{ik}} = \lambda u_{ll}\delta_{ik} + 2\mu u_{ik}.\]
    By defining \(\lambda = K - 2\mu/3\) we can write this as
    \[\sigma_{ik} = \left( K - \frac{2}{3}\mu \right)\delta_{ik}u_{ll} + 2\mu u_{ik}.\]
    We will see a physical interpretation for the constant \(K\) later.
    
    The equations above allow us to compute the stress tensor for a given deformation tensor assuming we know the values of the relevant constants.
    We would also like to be able to reverse this and work out what deformation caused a given stress.
    To this end consider the trace of the stress tensor:
    \[\sigma_{ll} = \left[ 3\left( K - \frac{2}{3}\mu \right) + 2\mu \right] u_{ll} = 3Ku_{ll}.\]
    Hence
    \[u_{ll} = \frac{1}{3K}\sigma_{ll}.\]
    Substituting this into the equation for \(\sigma_{ik}\) in terms of \(u\) and rearranging gives
    \[u_{ik} = \frac{1}{2\mu}\left[ \sigma_{ik} - \left( K - \frac{2}{3}\mu \right) \frac{1}{3K} \delta_{ik} \sigma_{ll} \right] = \frac{1}{2\mu}\sigma_{ik} - \frac{1}{6\mu}\delta_{ik}\sigma_{ll} + \frac{1}{9K}\delta_{ik}\sigma_{ll}.\]
    
    At this point we note that a given deformation can be split into two parts, one which characterises the shear part of the deformation and the other characterises the compression/extension part.
    The shear part is given by all of the off diagonal elements whereas the compression is given by the diagonal parts:
    \[u_{ik} = \underbrace{\left( u_{ik} - \frac{1}{3}\delta_{ik}u_{ll} \right)}_{\text{shear}} + \underbrace{\frac{1}{3}\delta_{ik}u_{ll}.}_{\text{compression}}\]
    Notice the factor of \(1/3\) which occurs since \(\delta_{ii} = 3\).
    We can then write the free energy as
    \[\freeEnergy = \mu\left( u_{ik} - \frac{1}{3}\delta_{ik}u_{ll} \right)^2 + \frac{K}{2}u_{ll}^2.\]
    This gives us a physical interpretation for \(K\) and \(\mu\).
    \(K\), called the \define{bulk modulus} or \define{compressability}, characterises how hard it is to compress a material.
    Similarly \(\mu\), called the \define{shear modulus}, characterises how hard it is to shear the material.
    
    We saw in example~\ref{exa:isotropic compression stress tensor} that a compression of force \(p\) per unit area leads to a stress tensor \(\sigma_{ik} = -p\delta_{ik}\).
    Hence the deformation tensor is
    \begin{align*}
        u_{ik} &=  \frac{1}{2\mu}\sigma_{ik} - \frac{1}{6\mu}\delta_{ik}\sigma_{ll} + \frac{1}{9K}\delta_{ik}\sigma_{ll}\\
        &= \frac{1}{2\mu}(-p\delta_{ik}) - \frac{1}{6\mu}\delta_{ik}(-p\delta_{ll}) + \frac{1}{9K}\delta_{ik}(-p\delta_{ll})\\
        &= -\frac{1}{2\mu}p\delta_{ik} + \frac{1}{2\mu}\delta_{ik}p - \frac{1}{3K}\delta_{ik}p\\
        &= - \frac{1}{3K}\delta_{ik}p.\\
    \end{align*}
    So
    \[
        u = 
        \begin{pmatrix}
            -p/3K & 0 & 0\\
            0 & -p/3K & 0\\
            0 & 0 & -p/3K
        \end{pmatrix}
        .
    \]
    Recall that the trace of the deformation tensor gives the relative volume change:
    \[u_{ll} = -\frac{p}{K} = \frac{\delta V}{V}.\]
    This shows that increasing \(K\) means that we need to increase the pressure to get the same volume change, which makes sense as \(K\) is the compressibility of the material.
    If we interpret this as a pressure change rather than an absolute pressure then we can rearrange to get
    \[\frac{1}{K} = -\frac{1}{V}\pdv{V}{P}\]
    which should be familiar from thermodynamics.
    In this case the change in free energy is given by
    \[\freeEnergy = \frac{p^2}{2K}\]
    where we have chosen to define zero free energy as the starting free energy of the system.
    Notice from this that \(K\), and by extension \(\mu\) and \(\lambda\) have units of energy per metre cubed which is the same as pascals.
    
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \tikzexternaldisable
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    
    \clearpage
    \appendix
    \part*{Appendix}
    \addcontentsline{toc}{part}{Appendix}
    \begingroup
    \let\clearpage\relax
    \input{appendices/appendices}
    \endgroup 
\end{document}
